---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}

    destfile <- file.path(".", "2025","activity.csv")
    data1 <- read.csv(destfile, colClasses = c("numeric","character","character"))
# total records
    dim(data1) 

```


    
## histogram of the total number of steps take each day

```{r}
    steps_per_day <- aggregate(steps~date, data = data1, sum, na.rm = TRUE)
    
    hist (steps_per_day$steps, main ="total steps per day", xlab = "steps", col="blue", breaks=20)
```

## What is mean total number of steps taken per day?

```{r}
    mean(steps_per_day$steps, na.rm = TRUE)
    
    median (steps_per_day$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?

```{r}
    avg_steps_interval <- aggregate(steps ~ interval, data = data1, mean, na.rm = TRUE)
    
    plot(avg_steps_interval$interval, avg_steps_interval$steps, 
     type = "l", 
     col = "blue", 
     lwd = 2,
     main = "Average Daily Activity Pattern",
     xlab = "5-Minute Interval", 
     ylab = "Average Steps Taken")


    max_interval <- avg_steps_interval[which.max(avg_steps_interval$steps), ]
    print(max_interval)
```

## Imputing missing values

# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA

```{r}
total_nas <- sum(is.na(data1$steps))
print(total_nas)
```

# 1. Average steps per interval
```{r}
avg_steps <- aggregate(steps ~ interval, data = data1, mean)
```


# 2. Copy of the dataset to keep the original data intact
```{r}
activity_imputed <- data1
```

# 3. The data must replace NAs
```{r}
for (i in 1:nrow(activity_imputed)) {
    if (is.na(activity_imputed$steps[i])) {
        # Find the mean for that specific interval
        interval_value <- activity_imputed$interval[i]
        steps_value <- avg_steps[avg_steps$interval == interval_value, "steps"]
        
        # Replace NA with the mean
        activity_imputed$steps[i] <- steps_value
    }
}


stepsByDayImputed <- aggregate(steps ~ date, data = activity_imputed, sum)

hist(stepsByDayImputed$steps, 
     main = "Total Steps Per Day (Imputed Data)", 
     xlab = "Steps", 
     col = "blue", 
     breaks = 20)
```

## Are there differences in activity patterns between weekdays and weekends?

# Convert interval to numeric
```{r}
activity_imputed$interval <- as.numeric(activity_imputed$interval)
```
    
# Date must be in Date format
```{r}
activity_imputed$date <- as.Date(activity_imputed$date)
```
    
# Factor variable with two levels: "weekday" and "weekend"
```{r}
activity_imputed$day_type <- ifelse(weekdays(activity_imputed$date) %in% c("Saturday","Sunday"),"weekend", "weekday")
activity_imputed$day_type <- as.factor(activity_imputed$day_type)
```
    
# Mean for each interval across day types
```{r}
avg_steps_daytype <- aggregate(steps ~ interval + day_type, data = activity_imputed, mean)
    
library(lattice)
xyplot(steps ~ interval | day_type, data = avg_steps_daytype, 
          type = "l", 
          layout = c(1, 2), 
          xlab = "Interval", 
          ylab = "Number of steps",
          main = "Weekday vs. Weekend Activity")

```